Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility

### 2025-11-21 paper/paper.tex
up antes de borrar tiempo

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility},
  pdfauthor={Andreas Laffert Tamayo},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\setstretch{1.5}
This document was last modified at 2025-11-17 19:08:04 and it was last
rendered at 2025-11-17 19:08:04.

@@

What is the legitimate extent of market inequality in the eyes of the
public? Since the early 1980s, many countries have experienced a
widespread retreat from universal welfare programs and a shift toward
the privatization and commodification of public goods, welfare policies,
and social services (\citeproc{ref-gingrich_making_2011}{Gingrich,
2011}; \citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin
America, as elsewhere, neoliberal reforms reshaped welfare-state
institutions by extending market logic into domains of social
reproduction that were traditionally governed by the state
(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). This transformation
reduced the role of public provision and increased the presence of
private actors in core social services
(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). Echoing Polanyi's
(\citeproc{ref-polanyi_great_1975}{1975}) insight that markets
constitute a distinct moral order, the institutional diffusion of market
rules has fostered a corresponding moral economy: a constellation of
norms and values concerning fair allocation, embedded in institutions
and shaping individual subjectivities
(\citeproc{ref-mau_inequality_2015}{Mau, 2015};
\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}). Within
this framework, a growing body of research examines the extent to which,
and the mechanisms by which, citizens consider it fair that the
allocation of services like health care, pensions, and education be
governed by market-based criteria---a phenomenon known as \emph{market
justice preferences} (\citeproc{ref-busemeyer_skills_2014}{Busemeyer,
2014}; \citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et
al., 2025}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
2024}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}).
Understanding these preferences is crucial, as they contribute to
legitimizing economic inequality by framing it as the fair result of
individual responsibility and limited state intervention
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}).

Existing literature shows that market justice preferences are shaped by
both the economic and institutional context and individuals' positions
within social stratification. Grounded in the notion that economic
institutions influence people's normative attitudes
(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998}), studies find
that countries with stronger public provision or more expansive welfare
states exhibit lower levels of market justice preferences
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020}), while
more privatized contexts show stronger support for market-based criteria
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such contexts, market
justice preferences tend to rise as individuals ``ascend'' the social
structure, with those in more privileged positions in terms of class,
education, and income holding stronger preferences for market-based
solutions compared to those in more disadvantaged or at-risk positions
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-otero_power_2024}{Otero \& Mendoza, 2024};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007};
\citeproc{ref-vondemknesebeck_are_2016}{Von Dem Knesebeck et al.,
2016}).

Market justice preferences are shaped not only by objective
socioeconomic conditions but also by popular beliefs about inequality.
Among these, meritocracy is a key normative principle underpinning
market-based distributive preferences
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Studies show that
individuals with stronger meritocratic beliefs tend to perceive less
inequality and legitimize it by attributing economic differences to
personal achievement (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
2023}; \citeproc{ref-mijs_paradox_2019}{Mijs, 2019};
\citeproc{ref-wilson_role_2003}{Wilson, 2003}). In highly unequal
societies where access to services is largely governed by market logic,
such beliefs play a critical role in normalizing inequality. Recent
evidence from Chile shows that students who believe effort and talent
are rewarded in their country express stronger preferences for
market-based access to healthcare, pensions, and education
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).

Although it is clear that one's social position influences market
justice preferences, the question of how upward or downward mobility
within the social structure affects these preferences remains
unanswered. This question is far from trivial, especially in Latin
America, where many have experienced various forms of mobility amid high
economic inequality and deep welfare privatization
(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023};
\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}; \citeproc{ref-torche_intergenerational_2014}{Torche, 2014}).
Social origins and destinations affect attitudes toward inequality in
distinct ways (\citeproc{ref-day_movin_2017}{Day \& Fiske, 2017};
\citeproc{ref-gugushvili_intergenerational_2016}{Gugushvili, 2016b},
\citeproc{ref-gugushvili_subjective_2017}{2017};
\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}; \citeproc{ref-mijs_belief_2022}{Mijs et al.,
2022}; \citeproc{ref-wen_does_2021}{Wen \& Witteveen, 2021}), while
movement between these positions exposes individuals to different
experiences and mechanisms that shape their views on what is fair
(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014};
\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Building on this
research, examining the effects of social mobility on market justice
preferences can help to illuminate how inequalities in access to social
services are justified among individuals who have experienced, or not,
changes in their social standing, and what are the normative mechanisms
that guide this justification (\citeproc{ref-mau_inequality_2015}{Mau,
2015}).

Beyond their isolated effects, social mobility and meritocratic beliefs
interact in complex ways to shape market justice preferences. Among
others, a key mechanism proposed in the literature to explain how
mobility influences distributive justice preferences is the
psychological process of self-serving bias in causal attribution
(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-schmidt_experience_2011}{Schmidt, 2011}). This bias
suggests that individuals attribute failures---such as downward
mobility---to external factors, while crediting successes---such as
upward mobility---to their own merit and effort
(\citeproc{ref-miller_selfserving_1975}{Miller \& Ross, 1975}). Those
who experience upward mobility tend to view their social position as
earned, making them more likely to believe that individuals are
responsible for their own success or failure. Research shows that upward
mobility is associated with weaker preferences for redistribution
(\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}; \citeproc{ref-schmidt_experience_2011}{Schmidt,
2011}) and stronger legitimacy of income inequality
(\citeproc{ref-shariff_income_2016}{Shariff et al., 2016}). In contrast,
individuals who experience downward mobility tend to blame structural
factors like inequality and are more supportive of redistribution while
rejecting merit-based explanations
(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014}). Taken
together, I argue that meritocratic beliefs may reinforce this
self-serving attribution mechanism by legitimizing one's social status
as the outcome of personal merit, closely tied to attribution bias.

Against this background, this article pursues two main objectives:
first, to analyze the extent to which intergenerational social mobility
influences market justice preferences regarding healthcare, pensions,
and education; and second, to examine how meritocratic beliefs may
moderate this relationship. Building on a theoretical framework that
emphasizes how neoliberal transformations---particularly through the
privatization and commodification of key areas of social
reproduction---have profoundly reshaped processes of subject formation
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}), the central argument is
that upward mobility increases support for market justice preferences,
while downward mobility decreases it. Moreover, meritocratic beliefs are
expected to moderate this relationship by reflecting a self-serving
attribution bias, whereby individuals justify their social position in
terms of personal merit.

This study focuses on Chile, a particularly intriguing case for
examining market justice preferences. Despite significant economic
growth and poverty reduction, Chile has some of the highest levels of
inequality in Latin America and among OECD countries
(\citeproc{ref-chancel_world_2022}{Chancel et al., 2022};
\citeproc{ref-flores_top_2020}{Flores et al., 2020}). This inequality
coexists with short-range upward mobility among lower-class segments
moving into middle strata, though strong barriers remain to reaching
higher positions (\citeproc{ref-espinoza_estratificacion_2013}{Espinoza
et al., 2013}; \citeproc{ref-torche_intergenerational_2014}{Torche,
2014}). What makes Chile especially salient is that much of this
inequality is rooted in deep neoliberal reforms that institutionalized
the privatization and commodification of key social sectors
(\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). Introduced
during the dictatorship (1973--1989) and expanded in democracy, these
reforms enabled the unprecedented emergence of markets in health,
pensions, and education, with provision segmented by individuals'
ability to pay and supported by public subsidies
(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}). In parallel---and
despite waves of protest against inequality and commodification from
2006 to 2019 (\citeproc{ref-somma_no_2021}{Somma et al.,
2021})---Chilean subjectivities have been increasingly shaped by
neoliberal discourses and market logics, influencing their attitudes
toward inequality and welfare distribution
(\citeproc{ref-araujo_desafios_2012}{Araujo \& Martuccelli, 2012};
\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}).

In this context, the questions that guide this research are as follows:

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\tightlist
\item
  To what extent does intergenerational social mobility influence market
  justice preferences regarding healthcare, pensions, and education in
  Chile?
\item
  How do meritocratic beliefs condition or moderate this relationship in
  the Chilean context?
\end{enumerate}

To address these questions, this study draws on large-scale,
representative survey data collected in 2018 from the urban Chilean
population aged 18 to 75 (n = 2,726). The next section outlines the
theoretical framework linking market justice preferences, social
mobility, and meritocratic beliefs, and proposes a set of hypotheses.
This is followed by a description of the data, variables, and analytical
strategy. The final sections present the empirical findings, offer an
interpretation of the results, and conclude with a discussion of their
implications.

@@

Meritocracy constitutes a central ideological framework for legitimizing
different types of social inequality, for instance through market
justice beliefs. Rooted in the belief that rewards and positions should
be allocated based on individual effort and talent, meritocracy operates
as a normative ideal and a descriptive belief about how society
functions. As initially conceptualized by Michael Young
(\citeproc{ref-young_rise_1958}{1958}), the term was meant to critique a
system in which merit-based stratification becomes a new form of
inequality. However, over time, meritocracy has been widely supported in
many societies as a fair and desirable principle of distribution,
particularly within liberal democracies and market-oriented economies
(\citeproc{ref-mijs_paradox_2019}{Mijs, 2019};
\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}). From a sociological
perspective, the belief in meritocracy is more than a cognitive
assessment; it reflects a moral lens through which individuals interpret
inequality. People who believe that success results from hard work and
talent are more likely to view social and economic disparities as
legitimate (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
\citeproc{ref-castillo_meritocracia_2019}{Castillo et al., 2019}).
Conversely, if they see outcomes as driven by luck, social origin, or
systemic barriers, inequality is more likely to be perceived as unjust.
This distinction becomes crucial in societies with persistent structural
inequality, where public narratives often emphasize personal
responsibility and merit while overlooking entrenched disadvantages.

@@

Despite recurrent social unrest, Chile presents a paradoxical
coexistence between strong conflict over inequality and widespread
public legitimation of it. The October 2019 ``social outburst'', a
period of mass protests and severe political repression throughout the
country, crystallized discontent over the privatization and
commodification of social services, alongside a crisis of political
legitimacy (\citeproc{ref-somma_no_2021}{Somma et al., 2021}). A survey
during the protests identified pensions, healthcare, and education as
the top demands
(\citeproc{ref-nucleodesociologiacontingente_informe_2020}{Núcleo de
Sociología Contingente, 2020}), reflecting dissatisfaction with Chile's
stratified welfare regime. Yet, this unrest coexists with strong
meritocratic beliefs and high tolerance for income inequality. Castillo
(\citeproc{ref-castillo_legitimacy_2011}{2011}) shows that wage gaps
between occupations are consistently justified, and the larger the
perceived gap, the greater its legitimation, an effect amplified by
Chile's structural inequality. Moreover, perceptions of unfairness due
to non-meritocratic factors often reinforce belief in effort and talent
as legitimate bases for success
(\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}),
which are associated with lower perceived injustice
(\citeproc{ref-castillo_meritocracia_2019}{Castillo et al., 2019}).
Mac-Clure et al. (\citeproc{ref-mac-clure_justicia_2024}{Mac-Clure et
al., 2024}) find that low and lower-middle status individuals are less
likely than upper-status groups to view educational income differences
as unfair. Qualitative research further reveals the internalization of
the enterprising self (\citeproc{ref-mau_inequality_2015}{Mau, 2015}),
understood as a way in which the subjectivization process arises in line
with positioning the self in the market, generating its own market value
and subjecting it to competition. Canales et al.
(\citeproc{ref-canalesceron_sujeto_2021}{2021}) describes how families
navigate school choice as investment decisions, while Panes
(\citeproc{ref-panes_criticas_2020}{2020}) finds that many workers frame
pension contributions as individual investments aimed at maximizing
future returns. These representations suggest that market-based moral
orientations are deeply embedded in Chilean subjectivities, offering
fertile ground to study how support for market justice varies across
welfare domains.

@@

@@

@@

@@

\bibitem[\citeproctext]{ref-castillo_perceptions_2025}
Castillo, J. C., Laffert, A., Carrasco, K., \& Iturra, J. (2025).
Perceptions of {Inequality} and {Meritocracy}: {Their Interplay} in
{Shaping Preferences} for {Market Justice} in {Chile} (2016-2023).
\emph{Under Review at Frontiers in Sociology}.

@@

@@

@@

@@

@@

@@

@@

>>>>

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility},
  pdfauthor={Andreas Laffert Tamayo},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\setstretch{1.5}
This document was last modified at 2025-11-21 16:32:59 and it was last
rendered at 2025-11-21 16:32:59.

@@

What is the legitimate scope of market inequality in the eyes of the
public? Since the early 1980s, many countries have experienced a
rollback of universal welfare programs and a shift toward the
privatisation and commodification of public goods and social services
(\citeproc{ref-gingrich_making_2011}{Gingrich, 2011};
\citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin America, these
reforms extended market logic into domains of social reproduction long
governed by the state, shrinking public provision and expanding private
actors in core welfare sectors (\citeproc{ref-ferre_welfare_2023}{Ferre,
2023}). From a moral-economy perspective, this diffusion of market rules
has reconfigured the normative order: through policy feedback, welfare
institutions increasingly embed market-based criteria of ``fair''
allocation and shape how citizens understand deservingness and the
balance between state and market (\citeproc{ref-koos_moral_2019}{Koos \&
Sachweh, 2019};
\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}).
Against this backdrop, growing scholarship examines \emph{market justice
preferences}---the extent to which citizens regard market-based criteria
as a fair basis for access to essential services such as healthcare,
education and pensions
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). These
orientations matter because they help legitimise economic inequality by
portraying stratified outcomes as a fair result of individual
responsibility and limited state intervention.

This article examines public support for the commodification of pension
welfare---specifically, whether people consider it fair that access to
better pensions depends on income or contributions. Pension
privatisation lies at the core of Latin America's trajectory of
commodification (\citeproc{ref-huber_political_2000}{Huber \& Stephens,
2000}), making pension systems a strategic domain for analysing how
market principles become normatively embedded
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). As many countries
expanded mandatory or voluntary individual-capitalisation schemes---most
radically in Chile---pensions became organised around contribution-based
entitlements and market performance
(\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
\citeproc{ref-verbic_political_2019}{Verbič \& Spruk, 2019}). This shift
reshapes not only how pension institutions allocate old-age risk,
heightening exposure to poverty in retirement and tying benefit levels
to lifetime labour-market trajectories
(\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019};
\citeproc{ref-oecd_pensions_2023}{OECD, 2023}), but also reshape the
justice principles citizens associate with old-age security: people have
to judge whether higher benefits should follow market-derived criteria
or solidaristic principles
(\citeproc{ref-arrizabalomontoro_debate_2019}{Arrizabalo Montoro et al.,
2019}; \citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012}). Although
we know something about what people consider a ``just'' monetary pension
(\citeproc{ref-castillo_deserving_2019}{Castillo, Olivos, et al., 2019})
and about support for public versus private provision
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020};
\citeproc{ref-jaime-castillo_public_2013}{Jaime-Castillo, 2013}), we
still lack systematic evidence on how citizens assess the fairness of
market-dependent access to pensions.

Research shows that support for market-based access to welfare services
is strongly stratified: individuals with higher income, education or
class position are more likely to endorse market-based criteria than
disadvantaged groups (\citeproc{ref-busemeyer_skills_2014}{Busemeyer,
2014}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). Evidence for
pensions points in the same direction: higher income and greater
financial gains in funded schemes correlate with stronger support for
market-oriented pension policies
(\citeproc{ref-kerner_pension_2020}{Kerner, 2020}). Yet we know very
little about how movement within the class structure shapes preferences
for pension commodification. This gap is especially salient in Latin
America, where mobility has unfolded under high inequality and deeply
privatised welfare regimes
(\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}). Social origins and destinations anchor material interests and
justice principles
(\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
\citeproc{ref-langsaether_explaining_2022}{Langsæther et al., 2022}),
while movement between these positions exposes individuals to new risks,
resources and normative environments that reshape fairness judgements
(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014};
\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023}). As Mau
(\citeproc{ref-mau_inequality_2015}{2015}) argues, social mobility can
foster market-justice orientations, helping explain how inequalities in
access to welfare services become normatively sustained.

The literature identifies several mechanisms through which mobility may
shape distributive-justice preferences---material self-interest
(\citeproc{ref-benabou_social_2001}{Benabou \& Ok, 2001}), socialisation
and acculturation
(\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}), learning through long-term exposure
(\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023}), and
cognitive attributions such as self-serving bias or privilege awareness
(\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-molina_its_2019}{Molina et al., 2019}). While distinct,
these accounts converge in highlighting two broad channels relevant for
market-based welfare: shifts in material stakes and shifts in normative
frames (\citeproc{ref-mau_inequality_2015}{Mau, 2015}). In this paper, I
do not treat these mechanisms as mediators but use them to guide
hypotheses about heterogeneity: if mobility alters preferences through
material interests and normative updating, its effects should be most
potent among those who interpret social outcomes through merit-based
evaluations of effort
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}; \citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}).

Against this background, this article pursues two aims: first, to assess
whether intergenerational mobility causally affects support for the
commodification of pensions; and second, to examine whether this effect
varies by meritocratic beliefs and over time. I argue that mobility
works through two main channels. Upward mobility increases earnings,
expected returns in a funded system and perceived economic security,
making contribution-based access more appealing. At the same time,
integration into higher-class networks exposes people to more
individualistic norms and effort-based understandings of fairness.
Downward mobility, in turn, heightens economic insecurity, dependence on
collective protection and attachment to solidaristic principles,
lowering support for market allocation. Meritocratic beliefs are
conceptualised not as mediators but as moderators. Among those who see
mobility as a fair reward for effort, both material and normative
effects should be amplified, producing higher support for pension
commodification. The paper contributes to research on market justice by
focusing on a specific welfare domain and by providing causal evidence
on how mobility shapes distributive preferences.

This study focuses on Chile, a particularly instructive case for
examining preferences for pension commodification. Despite sustained
economic growth, Chile combines high and persistent inequality with
short-range upward mobility and substantial barriers to higher-class
positions (\citeproc{ref-flores_top_2020}{Flores et al., 2020};
\citeproc{ref-salgado_uplifting_2025}{Salgado et al., 2025};
\citeproc{ref-torche_intergenerational_2014}{Torche, 2014}). It was the
first country worldwide to fully replace a public pay-as-you-go system
with a mandatory, privately administered (through pension fund
administrators, AFPs) defined-contribution scheme, while the state has
progressively expanded a segmented solidarity pillar to compensate for
insufficient benefits (\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
\citeproc{ref-madariaga_three_2020}{Madariaga, 2020};
\citeproc{ref-solimano_rise_2021}{Solimano, 2021}). In parallel---and
despite waves of protest from 2006 to 2019 targeting the AFP system
(\citeproc{ref-somma_no_2021}{Somma et al., 2021})---Chilean
subjectivities have been increasingly shaped by neoliberal discourses
and market logics, influencing attitudes toward pension distribution and
inequality (\citeproc{ref-araujo_desafios_2012}{Araujo \& Martuccelli,
2012}; \citeproc{ref-galvez_con_2023}{Gálvez, 2023}). The combination of
profound inequality, market-driven old-age security and contested
legitimacy makes Chile a critical case for analysing how mobility and
meritocratic beliefs structure support for pension commodification.

To address these questions, the study draws on large-scale,
representative survey data for the urban Chilean population collected in
2016, 2018 and 2023. It adopts a causal framework of mobility effects
inspired by Breen \& Ermisch (\citeproc{ref-breen_effects_2024}{2024}).
The following section outlines the theoretical framework linking
preferences for market-based welfare, social mobility and expected
heterogeneity by meritocratic beliefs and time, and derives hypotheses.
A subsequent section presents the causal identification strategy, the
data, the variables, and the analytical approach. The empirical sections
report the findings and interpret their implications. The article
concludes by discussing what these results reveal about the politics of
market-based pension welfare and social mobility.

@@

Meritocracy constitutes a central ideological framework for legitimizing
different types of social inequality, for instance through market
justice beliefs. Rooted in the belief that rewards and positions should
be allocated based on individual effort and talent, meritocracy operates
as a normative ideal and a descriptive belief about how society
functions. As initially conceptualized by Michael Young
(\citeproc{ref-young_rise_1958}{1958}), the term was meant to critique a
system in which merit-based stratification becomes a new form of
inequality. However, over time, meritocracy has been widely supported in
many societies as a fair and desirable principle of distribution,
particularly within liberal democracies and market-oriented economies
(\citeproc{ref-mijs_paradox_2019}{Mijs, 2019};
\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}). From a sociological
perspective, the belief in meritocracy is more than a cognitive
assessment; it reflects a moral lens through which individuals interpret
inequality. People who believe that success results from hard work and
talent are more likely to view social and economic disparities as
legitimate (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
\citeproc{ref-castillo_meritocracia_2019}{Castillo, Torres, et al.,
2019}). Conversely, if they see outcomes as driven by luck, social
origin, or systemic barriers, inequality is more likely to be perceived
as unjust. This distinction becomes crucial in societies with persistent
structural inequality, where public narratives often emphasize personal
responsibility and merit while overlooking entrenched disadvantages.

@@

Despite recurrent social unrest, Chile presents a paradoxical
coexistence between strong conflict over inequality and widespread
public legitimation of it. The October 2019 ``social outburst'', a
period of mass protests and severe political repression throughout the
country, crystallized discontent over the privatization and
commodification of social services, alongside a crisis of political
legitimacy (\citeproc{ref-somma_no_2021}{Somma et al., 2021}). A survey
during the protests identified pensions, healthcare, and education as
the top demands
(\citeproc{ref-nucleodesociologiacontingente_informe_2020}{Núcleo de
Sociología Contingente, 2020}), reflecting dissatisfaction with Chile's
stratified welfare regime. Yet, this unrest coexists with strong
meritocratic beliefs and high tolerance for income inequality. Castillo
(\citeproc{ref-castillo_legitimacy_2011}{2011}) shows that wage gaps
between occupations are consistently justified, and the larger the
perceived gap, the greater its legitimation, an effect amplified by
Chile's structural inequality. Moreover, perceptions of unfairness due
to non-meritocratic factors often reinforce belief in effort and talent
as legitimate bases for success
(\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}),
which are associated with lower perceived injustice
(\citeproc{ref-castillo_meritocracia_2019}{Castillo, Torres, et al.,
2019}). Mac-Clure et al.
(\citeproc{ref-mac-clure_justicia_2024}{Mac-Clure et al., 2024}) find
that low and lower-middle status individuals are less likely than
upper-status groups to view educational income differences as unfair.
Qualitative research further reveals the internalization of the
enterprising self (\citeproc{ref-mau_inequality_2015}{Mau, 2015}),
understood as a way in which the subjectivization process arises in line
with positioning the self in the market, generating its own market value
and subjecting it to competition. Canales et al.
(\citeproc{ref-canalesceron_sujeto_2021}{2021}) describes how families
navigate school choice as investment decisions, while Panes
(\citeproc{ref-panes_criticas_2020}{2020}) finds that many workers frame
pension contributions as individual investments aimed at maximizing
future returns. These representations suggest that market-based moral
orientations are deeply embedded in Chilean subjectivities, offering
fertile ground to study how support for market justice varies across
welfare domains.

@@

\bibitem[\citeproctext]{ref-arenas_sistemas_2019}
Arenas, A. (2019). \emph{{Los sistemas de pensiones en la encrucijada:
desafíos para la sostenibilidad en América Latina}}. Santiago: Naciones
Unidas, CEPAL.

@@

\bibitem[\citeproctext]{ref-arrizabalomontoro_debate_2019}
Arrizabalo Montoro, X., Del Rosal, M., \& Javier Murillo Arroyo, F.
(2019). The {Debate} on {Pension Systems}: {The Paradigmatic Cases} of
{Chile} and {Spain}. \emph{The American Journal of Economics and
Sociology}, \emph{78}(1), 195--223.
\url{https://doi.org/10.1111/ajes.12262}

@@

\bibitem[\citeproctext]{ref-borzutzky_pension_2012}
Borzutzky, S. (2012). Pension market failure in {Chile}: Foundations,
analysis and policy reforms. \emph{Journal of Comparative Social
Welfare}, \emph{28}(2), 103--112.
\url{https://doi.org/10.1080/17486831.2012.655980}

@@

\bibitem[\citeproctext]{ref-castillo_perceptions_2025}
Castillo, J. C., Laffert, A., Carrasco, K., \& Iturra-Sanhueza, J.
(2025). Perceptions of inequality and meritocracy: Their interplay in
shaping preferences for market justice in {Chile} (2016--2023).
\emph{Frontiers in Sociology}, \emph{10}, 1634219.
\url{https://doi.org/10.3389/fsoc.2025.1634219}

@@

\bibitem[\citeproctext]{ref-castillo_deserving_2019}
Castillo, J. C., Olivos, F., \& Azar, A. (2019). Deserving a {Just
Pension}: {A Factorial Survey Approach}*. \emph{Social Science
Quarterly}, \emph{100}(1), 359--378.
\url{https://doi.org/10.1111/ssqu.12539}

@@

\bibitem[\citeproctext]{ref-galvez_con_2023}
Gálvez, R. (2023). Con mi plata no\ldots{} alcanza. \emph{CNN Chile}.

@@

\bibitem[\citeproctext]{ref-huber_political_2000}
Huber, E., \& Stephens, J. D. (2000). \emph{The political economy of
pension reform: {Latin America} in comparative perspective}. Geneva:
UNRISD.

@@

\bibitem[\citeproctext]{ref-jaime-castillo_public_2013}
Jaime-Castillo, A. M. (2013). Public opinion and the reform of the
pension systems in {Europe}: The influence of solidarity principles.
\emph{Journal of European Social Policy}, \emph{23}(4), 390--405.
\url{https://doi.org/10.1177/0958928713507468}

@@

\bibitem[\citeproctext]{ref-kerner_pension_2020}
Kerner, A. (2020). Pension {Returns} and {Popular Support} for
{Neoliberalism} in {Post-Pension Reform Latin America}. \emph{British
Journal of Political Science}, \emph{50}(2), 585--620.
\url{https://doi.org/10.1017/S0007123417000710}

@@

\bibitem[\citeproctext]{ref-madero-cabib_private_2019}
Madero-Cabib, I., Biehl, A., Sehnbruch, K., Calvo, E., \& Bertranou, F.
(2019). Private {Pension Systems Built} on {Precarious Foundations}: {A
Cohort Study} of {Labor-Force Trajectories} in {Chile}. \emph{Research
on Aging}, \emph{41}(10), 961--987.
\url{https://doi.org/10.1177/0164027519874687}

@@

\bibitem[\citeproctext]{ref-oecd_pensions_2023}
OECD. (2023). \emph{Pensions at a {Glance} 2023: {OECD} and {G20
Indicators}}. OECD Publishing. \url{https://doi.org/10.1787/678055dd-en}
